---
title: "Lab Session – Multiple Regression in Econometrics"
subtitle: "Wage Determinants: An Application of the Mincer Equation"
author: "Student: ___________________________"
date: "Academic Year 2025-2026"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 7, fig.height = 4)
```

---

# Lab Overview

## Economic Context

The **Mincer equation** (1974) is one of the most widely estimated models in labor economics.
It relates the **logarithm of wages** to education, experience, and experience squared:

$$\ln(w_i) = \alpha + \beta_1 \cdot \text{educ}_i + \beta_2 \cdot \text{exp}_i + \beta_3 \cdot \text{exp}_i^2 + \varepsilon_i$$

This lab uses a simulated dataset of **500 employed individuals** and guides you through the
key steps of a rigorous OLS multiple regression analysis.

## Learning Objectives

By the end of this 2-hour lab, you will be able to:

1. Explore and visualize a dataset to motivate a regression model
2. Estimate and interpret a multiple OLS regression
3. Perform and interpret individual and joint hypothesis tests
4. Detect violations of Gauss-Markov assumptions (heteroskedasticity, multicollinearity)
5. Produce and interpret out-of-sample predictions

## Required Packages

```{r packages, eval=FALSE}
install.packages(c("tidyverse", "car", "lmtest", "sandwich",
                   "stargazer", "ggcorrplot", "broom"))
```

```{r load_packages}
library(tidyverse)
library(car)
library(lmtest)
library(sandwich)
library(stargazer)
library(ggcorrplot)
library(broom)
```

## Loading the Data

```{r load_data}
dataset <- read.csv("wages.csv", stringsAsFactors = FALSE)

# Convert categorical variables to factors
dataset <- dataset %>%
  mutate(
    city_size     = factor(city_size,     levels = c("rural", "small", "large")),
    contract_type = factor(contract_type, levels = c("fixed_term", "permanent", "self_employed"))
  )
```

### Variable Dictionary

| Variable        | Type        | Description                                           |
|:----------------|:------------|:------------------------------------------------------|
| `wage`          | Numeric     | Monthly net wage (USD)                                |
| `log_wage`      | Numeric     | Natural log of wage                                   |
| `education`     | Numeric     | Years of education (8–22)                             |
| `experience`    | Numeric     | Years of work experience (0–45)                       |
| `exp_squared`   | Numeric     | Experience² (Mincer quadratic term)                   |
| `age`           | Numeric     | Individual's age                                      |
| `female`        | Binary      | 1 = female, 0 = male                                  |
| `public_sector` | Binary      | 1 = public sector, 0 = private                        |
| `city_size`     | Categorical | "rural" (ref), "small", "large"                       |
| `contract_type` | Categorical | "fixed_term" (ref), "permanent", "self_employed"      |
| `num_children`  | Numeric     | Number of children                                    |
| `unionized`     | Binary      | 1 = union member, 0 = non-member                      |
| `hours`         | Numeric     | Weekly hours worked                                   |

---

# Part 1 – Data Exploration *(~20 min)*

## Descriptive Statistics

**Question 1.1** — Display the dimensions of the dataset and compute descriptive statistics
(mean, median, standard deviation, min, max) for all numeric variables.
How many men and women are in the sample? What is the breakdown by contract type?

```{r q1_1}
# Your code here

```

**Question 1.2** — Plot **two side-by-side histograms**: the distribution of raw `wage` and
that of `log_wage`. What do you observe about their shapes?
Justify the use of `log_wage` as the dependent variable in regression.

```{r q1_2, fig.width = 10}
# Your code here

```

## Correlations and Scatter Plots

**Question 1.3** — Compute and display the **correlation matrix** for the continuous numeric
variables (`log_wage`, `education`, `experience`, `age`, `hours`).
Identify strongly correlated pairs and discuss potential implications for the regression.

```{r q1_3, fig.width = 7, fig.height = 6}
# Your code here

```

**Question 1.4** — Create two scatter plots with a regression line:

- `log_wage` as a function of `education`
- `log_wage` as a function of `experience` (add a LOESS smooth curve in addition)

What do you observe about the shape of each relationship?

```{r q1_4, fig.width = 10}
# Your code here

```

---

# Part 2 – Simple OLS Regression *(~15 min)*

**Question 2.1** — Estimate the **simple regression** of `log_wage` on `education` alone.
Display the full `summary()` output. Interpret the coefficient on education in economic terms
(percentage return per additional year).

```{r q2_1}
# Your code here

```

**Question 2.2** — Compute **manually** (without `lm()`) the OLS estimators $\hat\beta_0$ and
$\hat\beta_1$ using:

$$\hat{\beta}_1 = \frac{\sum(x_i - \bar{x})(y_i - \bar{y})}{\sum(x_i - \bar{x})^2},
\qquad \hat{\beta}_0 = \bar{y} - \hat{\beta}_1\bar{x}$$

Verify that your results match those from `lm()`.

```{r q2_2}
# Your code here

```

---

# Part 3 – Multiple OLS Regression *(~35 min)*

## The Standard Mincer Model

**Question 3.1** — Estimate the **standard Mincer model**:

$$\ln(w_i) = \beta_0 + \beta_1\,\text{educ}_i + \beta_2\,\text{exp}_i + \beta_3\,\text{exp}^2_i + \varepsilon_i$$

Display the summary. How do the $R^2$ and the coefficient on education change compared
to the simple regression? Interpret every coefficient.

```{r q3_1}
# Your code here

```

## Extended Model

**Question 3.2** — Estimate an **extended model** by adding: `female`, `public_sector`,
`city_size`, `contract_type`, `unionized`, `num_children`.
Interpret the coefficients on the binary variables and the reference categories.

```{r q3_2}
# Your code here

```

**Question 3.3** — Using the extended model, answer the following:

a. What is the estimated wage premium (in %) for working in the public sector?
b. What is the estimated gender wage gap (in %)?
c. What is the marginal effect of one additional year of experience at 10 years? At 30 years?
   Use: $\dfrac{\partial \ln(w)}{\partial \text{exp}} = \hat\beta_2 + 2\hat\beta_3 \cdot \text{exp}$

```{r q3_3}
# Your code here

```

## Comparative Regression Table

**Question 3.4** — Produce a **regression table** comparing all three models (simple, Mincer,
extended) using `stargazer()`. Compare the adjusted $R^2$ values.
Is the added complexity of the extended model justified?

```{r q3_4, results='asis'}
# Your code here

```

---

# Part 4 – Hypothesis Testing *(~25 min)*

## Individual t-tests

**Question 4.1** — From the extended model, identify which coefficients are **not significant**
at the 5% level. Report the $t$-statistic and p-value for `unionized`.
State the null hypothesis tested by the t-test.

```{r q4_1}
# Your code here

```

## Global F-test

**Question 4.2** — Interpret the **F-statistic** reported at the bottom of the extended model
summary. What is its null hypothesis? What do you conclude?

```{r q4_2}
# Your code here

```

## Joint Restrictions (linearHypothesis)

**Question 4.3** — Test the joint hypothesis that **experience has no effect on wages**:
$$H_0:\;\beta_{\text{exp}} = 0 \;\text{ and }\; \beta_{\text{exp}^2} = 0$$
Use `linearHypothesis()` from the `car` package. State your conclusion.

```{r q4_3}
# Your code here

```

**Question 4.4** — An economist claims the return to education is **exactly 8%**
($H_0: \beta_{\text{educ}} = 0.08$). Test this with `linearHypothesis()`.
Is the value 0.08 compatible with the data at the 5% level?

```{r q4_4}
# Your code here

```

---

# Part 5 – Gauss-Markov Diagnostics *(~20 min)*

## Heteroskedasticity

**Question 5.1** — Run the **Breusch-Pagan test** (`bptest()` from `lmtest`).
State $H_0$ and $H_1$ clearly. Conclude at the 5% level.

```{r q5_1}
# Your code here

```

**Question 5.2** — Re-estimate the extended model with **HC3 heteroskedasticity-robust
standard errors** (`coeftest()` + `vcovHC()`). Compare the standard errors and p-values
before and after correction. Do your conclusions change?

```{r q5_2}
# Your code here

```

## Multicollinearity

**Question 5.3** — Compute the **Variance Inflation Factor (VIF)** for each variable in the
extended model using `vif()`. Which thresholds are commonly used?
Is there a multicollinearity problem here?

```{r q5_3}
# Your code here

```

---

# Part 6 – Prediction *(~5 min)*

**Question 6.1** — Using the extended model, predict the **monthly wage** (in USD, not log)
for the two profiles below. Report 95% **confidence intervals** (for the mean prediction)
and **prediction intervals** (for an individual). Explain the difference between the two.

| Profile   | educ | exp | female | public_sector | city_size | contract_type | num_children | unionized | hours |
|:----------|:----:|:---:|:------:|:-------------:|:---------:|:-------------:|:------------:|:---------:|:-----:|
| Profile A | 17   | 10  | 0      | 0             | large     | permanent     | 0            | 0         | 39    |
| Profile B | 12   | 20  | 1      | 1             | rural     | fixed_term    | 2            | 0         | 35    |

```{r q6_1}
# Your code here

```

---

# Summary

**Final question** — Write a short paragraph (8–12 lines) summarising your main econometric
findings as you would in an academic paper. Address:

- The main determinants of wages identified
- Robustness of the results (tests performed)
- Limitations of the analysis (omitted variables, causality, etc.)

> **[Write your summary here]**

---

# Grading Rubric

| Section | Points |
|:--------|:------:|
| 1 – Data Exploration | 15 |
| 2 – Simple Regression | 10 |
| 3 – Multiple Regression | 25 |
| 4 – Hypothesis Testing | 25 |
| 5 – Gauss-Markov Diagnostics | 15 |
| 6 – Prediction | 10 |
| **Total** | **100** |

---

*Duration: 2 hours — All materials allowed — R and RStudio required*
