---
title: "Lab Session – Multiple Regression in Econometrics"
subtitle: "SOLUTION TEMPLATE — Fill in the blanks"
author: "Student: ___________________________"
date: "Academic Year 2025-2026"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 7, fig.height = 4)
```

> **Instructions for this template:**
> Each code chunk contains scaffolding comments (lines starting with `#`) that guide you
> step by step. Replace every `___` with appropriate R code.
> Write your textual interpretations in the `> **Interpretation:**` blocks.

---

# Setup

```{r load_packages}
# Load required packages (install first if needed)
library(tidyverse)
library(car)        # linearHypothesis(), vif()
library(lmtest)     # bptest()
library(sandwich)   # vcovHC()
library(stargazer)  # regression tables
library(ggcorrplot) # correlation matrix plot
library(broom)      # tidy()
```

```{r load_data}
# Load the dataset
dataset <- read.csv("wages.csv", stringsAsFactors = FALSE)

# Convert categorical variables to ordered factors
# Reference levels: city_size = "rural", contract_type = "fixed_term"
dataset <- dataset %>%
  mutate(
    city_size     = factor(city_size,     levels = c("rural", "small", "large")),
    contract_type = factor(contract_type, levels = c("fixed_term", "permanent", "self_employed"))
  )
```

---

# Part 1 – Data Exploration *(~20 min)*

## Q1.1 — Descriptive Statistics

```{r q1_1}
# 1. Print the dataset dimensions
dim(___)

# 2. Display variable types using glimpse
glimpse(___)

# 3. Summary statistics for numeric variables
# Hint: use summary() or summarise(across(...))
num_vars <- c("wage", "log_wage", "education", "experience",
              "age", "hours", "num_children")

dataset %>%
  select(all_of(___)) %>%
  summary()

# 4. Count men and women (female = 0 or 1)
table(dataset$___)

# 5. Contract type breakdown
table(dataset$___)
```

> **Interpretation:**
> [Describe the main characteristics of the sample: average wage, education level,
>  gender balance, contract type distribution.]

## Q1.2 — Wage Distributions

```{r q1_2, fig.width = 10, fig.height = 4}
# Hint: use ggplot() + geom_histogram()
# Create two histograms side by side:
#   - Left panel: raw 'wage' variable
#   - Right panel: 'log_wage' variable
# Use gridExtra::grid.arrange(p1, p2, ncol = 2) to arrange them

p1 <- ggplot(dataset, aes(x = ___)) +
  geom_histogram(bins = 40, fill = "#2196F3", color = "white", alpha = 0.8) +
  geom_vline(xintercept = mean(dataset$wage),   color = "red",    linetype = "dashed") +
  geom_vline(xintercept = median(dataset$wage), color = "orange", linetype = "dashed") +
  labs(title = "Raw Wage", subtitle = "Red = mean | Orange = median",
       x = "Wage (USD)", y = "Count") +
  theme_minimal()

p2 <- ggplot(dataset, aes(x = ___)) +
  geom_histogram(bins = 40, fill = "#4CAF50", color = "white", alpha = 0.8) +
  labs(title = "Log-Wage",
       x = "Log-Wage", y = "Count") +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

> **Interpretation:**
> [Describe the shapes of both distributions. Why is log_wage better suited for regression?
>  Give at least 2 reasons.]

## Q1.3 — Correlation Matrix

```{r q1_3, fig.width = 7, fig.height = 6}
# Select the continuous variables and compute the correlation matrix
cor_vars <- dataset %>%
  select(___, ___, ___, ___, ___)   # log_wage, education, experience, age, hours

mat_cor <- cor(___, use = "complete.obs")
round(mat_cor, 3)

# Plot with ggcorrplot
ggcorrplot(mat_cor,
           method = "circle", type = "lower",
           lab = TRUE, lab_size = 3,
           colors = c("#E91E63", "white", "#2196F3"),
           title = "Correlation Matrix") +
  theme_minimal()
```

> **Interpretation:**
> [Identify the two most strongly correlated pairs. What does this imply for the regression?]

## Q1.4 — Scatter Plots

```{r q1_4, fig.width = 10, fig.height = 4}
# Scatter plot 1: log_wage ~ education with linear fit
p1 <- ggplot(dataset, aes(x = ___, y = ___)) +
  geom_jitter(alpha = 0.3, color = "#2196F3", width = 0.2) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Log-Wage vs Education",
       x = "Years of Education", y = "Log-Wage") +
  theme_minimal()

# Scatter plot 2: log_wage ~ experience with both linear and LOESS fits
p2 <- ggplot(dataset, aes(x = ___, y = ___)) +
  geom_jitter(alpha = 0.3, color = "#4CAF50", width = 0.3) +
  geom_smooth(method = "lm",    color = "red",       se = TRUE, linetype = "dashed") +
  geom_smooth(method = "loess", color = "darkgreen", se = TRUE) +
  labs(title = "Log-Wage vs Experience",
       subtitle = "Red (dashed) = linear | Green = LOESS",
       x = "Years of Experience", y = "Log-Wage") +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

> **Interpretation:**
> [Describe the shape of each relationship. Why does the experience-wage plot motivate
>  including an experience-squared term?]

---

# Part 2 – Simple OLS Regression *(~15 min)*

## Q2.1 — Simple Regression

```{r q2_1}
# Estimate: log_wage ~ education
model_simple <- lm(___ ~ ___, data = dataset)

# Display the full summary
summary(___)
```

> **Interpretation:**
> - Coefficient on education: ___
> - Economic meaning (percentage return): ___
> - R²: ___ — meaning: ___

## Q2.2 — Manual OLS Computation

```{r q2_2}
x <- dataset$education
y <- dataset$log_wage

# Compute beta_1 manually using the OLS formula:
# beta1 = sum((x - x_bar)(y - y_bar)) / sum((x - x_bar)^2)
beta1_manual <- sum((___ - mean(___)) * (___ - mean(___))) /
                sum((___ - mean(___))^2)

# Compute beta_0:
# beta0 = y_bar - beta1 * x_bar
beta0_manual <- mean(___) - ___ * mean(___)

cat("Manual β̂₀:", round(beta0_manual, 6), "\n")
cat("Manual β̂₁:", round(beta1_manual, 6), "\n")

cat("lm() β̂₀  :", round(coef(model_simple)[1], 6), "\n")
cat("lm() β̂₁  :", round(coef(model_simple)[2], 6), "\n")

# Check: differences should be ~0
cat("Difference β₀:", abs(beta0_manual - coef(model_simple)[1]), "\n")
cat("Difference β₁:", abs(beta1_manual - coef(model_simple)[2]), "\n")
```

> **Comment:**
> [Are the results identical? What does this confirm about the formula?]

---

# Part 3 – Multiple OLS Regression *(~35 min)*

## Q3.1 — Standard Mincer Model

```{r q3_1}
# Estimate the Mincer model: log_wage ~ education + experience + exp_squared
model_mincer <- lm(___ ~ ___ + ___ + ___, data = dataset)
summary(___)
```

> **Interpretation:**
> - Coefficient on education: ___ (compare to simple model: ___)
> - Coefficient on experience: ___ (sign: ___, meaning: ___)
> - Coefficient on exp_squared: ___ (sign: ___, meaning: ___)
> - Adjusted R² improved from ___ to ___

## Q3.2 — Extended Model

```{r q3_2}
# Add: female, public_sector, city_size, contract_type, unionized, num_children
model_extended <- lm(log_wage ~ education + experience + exp_squared +
                       ___ + ___ + ___ + ___ +
                       ___ + ___,
                     data = dataset)
summary(___)
```

> **Interpretation:**
> - `female` coefficient: ___ → women earn approximately ___% [more/less] than men
> - `public_sector` coefficient: ___ → public sector premium ≈ ___%
> - Reference for `city_size`: "rural" — "small" coefficient: ___, "large": ___
> - Reference for `contract_type`: "fixed_term" — "permanent" coefficient: ___

## Q3.3 — Economic Calculations

```{r q3_3}
coefs <- coef(model_extended)

# a) Public sector wage premium (%)
# Hint: for a dummy variable, % effect = (exp(beta) - 1) * 100
premium_pub <- (exp(coefs[___]) - 1) * 100
cat("a) Public sector premium:", round(premium_pub, 2), "%\n")

# b) Gender wage gap (%)
gap_gender <- (exp(coefs[___]) - 1) * 100
cat("b) Gender wage gap:", round(gap_gender, 2), "%\n")

# c) Marginal return to experience
# Formula: d ln(w)/d exp = beta_exp + 2 * beta_exp2 * exp
beta_exp  <- coefs["experience"]
beta_exp2 <- coefs["exp_squared"]

# At 10 years:
marginal_10 <- (___ + 2 * ___ * 10) * 100
cat("c) Marginal return at 10 years:", round(marginal_10, 3), "%\n")

# At 30 years:
marginal_30 <- (___ + 2 * ___ * ___) * 100
cat("   Marginal return at 30 years:", round(marginal_30, 3), "%\n")

# Peak experience: set marginal return = 0 and solve
# 0 = beta_exp + 2 * beta_exp2 * exp* --> exp* = -beta_exp / (2 * beta_exp2)
exp_peak <- -___ / (2 * ___)
cat("   Wages peak at:", round(exp_peak, 1), "years of experience\n")
```

## Q3.4 — Comparative Regression Table

```{r q3_4, results='asis'}
stargazer(model_simple, model_mincer, model_extended,
          type = "html",
          title = "Table 1 – OLS Estimates of Log-Wage",
          column.labels = c("(1) Simple", "(2) Mincer", "(3) Extended"),
          dep.var.label = "Log-Wage",
          omit.stat = c("f", "ser"),
          digits = 4,
          notes = "Standard errors in parentheses. *** p<0.01, ** p<0.05, * p<0.1")
```

> **Comment:**
> [Compare the adjusted R² values across models. Is the added complexity justified?]

---

# Part 4 – Hypothesis Testing *(~25 min)*

## Q4.1 — Individual t-tests

```{r q4_1}
# Use broom::tidy() to get a clean coefficient table
coef_table <- tidy(model_extended) %>%
  select(term, estimate, std.error, statistic, p.value) %>%
  mutate(across(where(is.numeric), ~round(., 4)),
         significant_5pct = p.value < ___)     # fill in the threshold

print(coef_table, n = 20)

# Report details for 'unionized'
cat("\nFor 'unionized':\n")
cat("t-statistic:", coef_table$statistic[coef_table$term == "unionized"], "\n")
cat("p-value    :", coef_table$p.value[coef_table$term   == "unionized"], "\n")
```

> **H₀ tested by the t-test:** H₀: β_k = ___ vs H₁: β_k ≠ ___
>
> **Conclusion for `unionized`:** [Is the coefficient significant at 5%? What does this mean?]

## Q4.2 — Global F-test

```{r q4_2}
# Extract the F-statistic from the model summary
f_stat <- summary(model_extended)$fstatistic
cat("F-statistic:", round(f_stat[1], 2), "\n")
cat("df:", f_stat[2], "and", f_stat[3], "\n")

# Compute the p-value
p_val_F <- pf(f_stat[1], f_stat[2], f_stat[3], lower.tail = FALSE)
cat("p-value:", p_val_F, "\n")
```

> **H₀ of the F-test:** H₀: β₁ = β₂ = ⋯ = β_K = ___
>
> **Conclusion:** [Is the model globally significant? What does this test tell us?]

## Q4.3 — Joint Test: Experience

```{r q4_3}
# Test H0: beta_experience = 0 AND beta_exp_squared = 0 jointly
test_exp <- linearHypothesis(model_extended,
                             c("experience = 0", "exp_squared = 0"))
print(___)
```

> **Conclusion:** [Report the F-statistic and p-value. Do we reject H₀? What does this imply?]

## Q4.4 — Test: Return to Education = 8%

```{r q4_4}
# Test H0: beta_education = 0.08
test_educ <- linearHypothesis(model_extended, "education = ___")
print(___)

# Also show the 95% confidence interval for the education coefficient
confint(model_extended)["education", ]
```

> **Conclusion:** [Is 0.08 within the 95% confidence interval? Do we reject H₀ at 5%?]

---

# Part 5 – Gauss-Markov Diagnostics *(~20 min)*

## Q5.1 — Breusch-Pagan Test

```{r q5_1}
# Run the Breusch-Pagan test on the extended model
test_bp <- bptest(___)
print(test_bp)

# State your conclusion
cat("\nH₀:", "___\n")
cat("H₁:", "___\n\n")
cat("Conclusion at 5%:", ifelse(test_bp$p.value < 0.05, "Reject H₀", "Do not reject H₀"), "\n")
```

> **Consequence if H₀ is rejected:**
> [What happens to OLS estimators under heteroskedasticity? Are they still BLUE?]

## Q5.2 — Robust Standard Errors

```{r q5_2}
# Classic standard errors
cat("=== Classic SEs ===\n")
coeftest(___) %>% round(4)

# HC3 Robust standard errors
cat("\n=== HC3 Robust SEs ===\n")
robust_result <- coeftest(___, vcov = vcovHC(___, type = "HC3"))
print(round(robust_result, 4))

# Build a comparison table
se_classic <- sqrt(diag(vcov(___)))
se_robust  <- sqrt(diag(vcovHC(___, type = "HC3")))

data.frame(
  variable   = names(se_classic),
  SE_classic = round(se_classic, 5),
  SE_robust  = round(se_robust,  5),
  ratio      = round(se_robust / se_classic, 3)
) %>% arrange(desc(ratio)) %>% print()
```

> **Comparison:**
> [Which coefficients have the largest ratio? Do significance conclusions change?]

## Q5.3 — Variance Inflation Factors

```{r q5_3}
# Compute VIF for the extended model
vif_vals <- vif(___)
print(vif_vals)

cat("\nThresholds: VIF < 5 = OK | 5–10 = moderate | > 10 = severe\n")
```

> **Conclusion:**
> [Are there any VIF values above 5? Is multicollinearity a problem in this model?]

---

# Part 6 – Prediction *(~5 min)*

## Q6.1 — Predicting Wages for Two Profiles

```{r q6_1}
# Create the new data frame for prediction
new_profiles <- data.frame(
  education     = c(___, 12),   # Profile A = 17
  experience    = c(10, ___),   # Profile B = 20
  exp_squared   = c(___, ___),  # = experience^2
  female        = c(___, 1),    # Profile A = male (0)
  public_sector = c(0, ___),    # Profile B = public (1)
  city_size     = factor(c("large", "rural"), levels = c("rural", "small", "large")),
  contract_type = factor(c("permanent", "fixed_term"),
                         levels = c("fixed_term", "permanent", "self_employed")),
  num_children  = c(0, ___),    # Profile B = 2
  unionized     = c(0, 0),
  hours         = c(39, 35)
)
rownames(new_profiles) <- c("Profile A", "Profile B")

# 95% Confidence Intervals (for the mean prediction)
pred_ci <- predict(model_extended, newdata = new_profiles,
                   interval = "___", level = 0.95)

# 95% Prediction Intervals (for an individual)
pred_pi <- predict(model_extended, newdata = new_profiles,
                   interval = "___", level = 0.95)

# Convert from log to USD
results <- data.frame(
  Profile             = rownames(new_profiles),
  Log_Wage_Pred       = round(pred_ci[, "fit"], 4),
  Wage_Median         = round(exp(pred_ci[, "fit"])),
  CI_Lower            = round(exp(pred_ci[, "lwr"])),
  CI_Upper            = round(exp(pred_ci[, "upr"])),
  PI_Lower            = round(exp(pred_pi[, "lwr"])),
  PI_Upper            = round(exp(pred_pi[, "upr"]))
)
print(results)
```

> **Interpretation:**
> - Profile A predicted wage: ___. Profile B predicted wage: ___.
>
> **CI vs PI:**
> - Confidence interval: uncertainty about ___.
> - Prediction interval: uncertainty about ___. Always wider because ___.

---

# Summary

```{r summary_chunk}
# Optional: any additional calculations for your summary paragraph
cat("Adjusted R² (extended model):", round(summary(model_extended)$adj.r.squared, 4), "\n")
```

> **[Write your 8–12 line academic summary here]**
>
> Include: main wage determinants identified | tests performed and their outcomes |
> limitations (omitted variables, causality, data quality).

---

# Grading Rubric

| Section | Points |
|:--------|:------:|
| 1 – Data Exploration | 15 |
| 2 – Simple Regression | 10 |
| 3 – Multiple Regression | 25 |
| 4 – Hypothesis Testing | 25 |
| 5 – Gauss-Markov Diagnostics | 15 |
| 6 – Prediction | 10 |
| **Total** | **100** |

---

*Duration: 2 hours — All materials allowed — R and RStudio required*
